When Do You Repeat Yourself?\\ Voices from the Trenches of Linux Kernel Maintainers on Code Duplication

### 2025-10-29 papers/techdebt2026/techdebt.tex
xtechdebt2026: minor adjustments
Signed-off-by: Paulo Meirelles <paulormm@ime.usp.br>

We executed the tool in the AMD Display driver codebase. We manually analyzed the biggest duplications regarding the number of lines, finding one pair that we judged promising to try to deduplicate. We approached big duplications as proof of concept to gauge the tool's potential for enabling significant impacts.

Given the duplicated function pair, we observed that the files in the functions contained multiple other duplications. Thus, we proposed a simple systematic approach to mitigate all the functions duplicated in the context, not just the duplicated function pairs. After approaching the duplications with a systematic strategy, we sent a patch to the AMD development mailing list for the feedback of maintainers, while documenting the process, interactions, and impressions (available in our Replication Package).

@@

It is important to stress that, \textbf{no matter the chosen approach}, we supported all groups by providing review cycles on their contributions before submission, to prepare students for the interactions with the maintainers and alleviate rough, but natural, newcomers' mistakes. After the submission, we helped students understand the feedback from maintainers and gave directions on how to move forward. This close, yet indirect, participation through intimate mentoring characterizes the second phase as our participant-as-observer observation.

For this second phase, our primary data source was the mailing list threads, where all development occurred (available in our Replication Package). Using blog posts, the groups documented their experiences submitting a patch to the Linux kernel~[omitted], and we also leveraged those to complement our observations. % (available in our Replication Package).
%TODO: Add blogs ao pacote de replicação

\section{Results}

@@

We derived four takeaways from our \textbf{participant-as-observer observation}. We substantiate these takeaways with quotes from the maintainers that originated during the review process~\footnote{We \textbf{do not claim} these quotes fully represent the maintainers’ stance on code duplication.}.

@@

\begin{highlight-box}{Takeaway T1 (Readability)}
Maintainers prioritized readability over removing duplications in many different contexts. In their view, going against the \textbf{DRY principle} was worth it if it resulted in easier-to-read and understand code.
\end{highlight-box}

@@

Although removing duplications does not necessarily mean reducing lines of code, the patch from Group 6 added more net lines, which the maintainer pointed out as a drawback. Still, the real reason for the rejection was that the deduplication scattered parts of a complex code that resided close together before: \textit{``Wrapping this up doesn't provide any real advantage, requiring as it does the reviewer to look at this function AND where the value is set rather than seeing them in one place.''}. Needing to jump to a function definition (to see its behavior) or a function call (to see its concrete parameters) in order to understand code is a cognitively demanding task for developers~\cite{questions-programmers-ask}. Merging these informations, we derived our third takeaway:

@@

The reason for rejecting the patch from Group 10 involved the deduplication lowering readability, as the maintainer was \textit{``[...] not sure the code reduction is sufficiently to cover the resulting loss of readability.''}, reinforcing takeaway T1.

@@

Groups 7 and 8 used the extract method using helpers to remove duplicated code in the function's body. Group 7 refined its patch from the maintainers' feedback, resulting in a smooth acceptance process that did not produce any insights. Group 8 followed a similar course, but generated an interaction reinforcing takeaway T4. In the feedback of version 1, a maintainer claimed that \textit{``Even though there is less code repetition, we now have an extra comparison [...]''}, with another maintainer corroborating this in version three by saying \textit{``There is always a balance/trade-off between modularity and execution speed. I agree with \texttt{anonymous-maintainer}'s reply in the first patch [...]''}, regardless of the fact that the patch was eventually accepted in version four.

@@

Overall, these findings reveal that, within the Linux kernel community, the evaluation of code duplication extends beyond technical redundancy. Maintainability, clarity, and practical trade-offs play a decisive role in shaping maintainers’ judgments, suggesting that DRY is sometimes less valuable than keeping the code understandable, efficient, and integrated smoothly into a large and evolving system.

@@

\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{IEEEabrv,references}

>>>>

We executed the tool in the AMD Display driver codebase and manually analyzed the biggest duplications regarding the number of lines, finding one pair that we judged promising to try to deduplicate. We approached big duplications as proof of concept to gauge the tool's potential for enabling significant impacts.

Given the duplicated function pair, we observed that the files in the functions contained multiple other duplications. Thus, we proposed a simple systematic approach to mitigate all the functions duplicated in the context, not just the duplicated function pairs. After approaching the duplications with a systematic strategy, we sent a patch to the AMD development mailing list for the feedback of maintainers, while documenting the process, interactions, and impressions (available in our replication package\footnote{\url{https://doi.org/10.5281/zenodo.17432193}}	).

@@

It is important to stress that, \textbf{no matter the chosen approach, we supported all groups by providing review cycles on their contributions before submission}, to prepare students for the interactions with the maintainers and alleviate rough, but natural, newcomers' mistakes. After the submission, we helped students understand the feedback from maintainers and gave directions on how to move forward. This close, yet indirect, participation through intimate mentoring characterizes the second phase as our participant-as-observer observation.

For this second phase, our primary data source was the mailing list threads, where all development occurred (also available in our replication package). Using blog posts, the groups documented their experiences submitting a patch to the Linux kernel~[omitted], and we also leveraged those to complement our observations. % (available in our Replication Package).
%TODO: Add blogs ao pacote de replicação

\vspace{-.2cm}
\section{Results}

@@

We derived four takeaways from our \textbf{participant-as-observer observation}. We substantiate these takeaways with quotes from the maintainers that originated during the review process\footnote{We \textbf{do not claim} these quotes fully represent the maintainers’ stance on code duplication.}.

@@

\begin{highlight-box}{Takeaway T1 (Readability)}
Maintainers prioritized readability over removing duplications in many different contexts. In their view, going against the DRY principle was worth it if it resulted in easier-to-read and understand code.
\end{highlight-box}

@@

Although removing duplications does not necessarily mean reducing lines of code, the patch from Group 6 resulted in a positive net change in lines of code, which the maintainer pointed out as a drawback. Still, the real reason for the rejection was that the deduplication scattered parts of a complex code that resided close together before: \textit{``Wrapping this up doesn't provide any real advantage, requiring as it does the reviewer to look at this function AND where the value is set rather than seeing them in one place.''}. Needing to jump to a function definition (to see its behavior) or a function call (to see its concrete parameters) to understand code is a cognitively demanding task for developers~\cite{questions-programmers-ask}. Merging these informations, we derived our third takeaway:

@@

The reason for rejecting the patch from Group 10 involved the deduplication lowering readability, as the maintainer was \textit{``[...] not sure the code reduction is sufficiently to cover the resulting loss of readability.''}, reinforcing the takeaway T1.

@@

Groups 7 and 8 used the extract method using helpers to remove duplicated code in the function's body. Group 7 refined its patch from the maintainers' feedback, resulting in a smooth acceptance process that did not produce any insights. Group 8 followed a similar course, but generated an interaction reinforcing the takeaway T4. In the feedback of version 1, a maintainer claimed that \textit{``Even though there is less code repetition, we now have an extra comparison [...]''}, with another maintainer corroborating this in version three by saying: \textit{``There is always a balance/trade-off between modularity and execution speed. I agree with \texttt{anonymous-maintainer}'s reply in the first patch [...]''}, regardless of the fact that the patch was eventually accepted in version four.

@@

Overall, these findings reveal that, within the Linux kernel community, the evaluation of code duplication extends beyond technical redundancy. Maintainability, clarity, and practical trade-offs play a decisive role in shaping maintainers’ judgments, suggesting that the DRY principle is sometimes less valuable than keeping the code understandable, efficient, and integrated smoothly into a large and evolving system.

@@

\newpage
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{IEEEabrv,references}
